---
title: "Kiwifruit_Lab"
author: "Me :)"
date: "2026-07-22"
output: html_document
---

If you do not have the required packages run the code block below (you will need to remove the comment (\#))

```{r}
#install.packages("BiocManager")
#BiocManager::install(c("DESeq2", "clusterProfiler", "GO.db", "stringr"))
#install.packages(c("dplyr", "ggplot2", "pheatmap", "RColorBrewer", "tidyr", "readr", "tibble", "knitr", "kableExtra"))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(tidyr)
library(clusterProfiler)
library(readr)
library(GO.db)
library(tibble)
library(knitr)
library(kableExtra)
library(stringr)
```

# Identifying variants implicated in kiwifruit actinidain production

## The Purpose and the Plan

Using publicly available RNA-seq data from three different strains of kiwifruit (Hongyang (Red), G3 (Yellow) and Lvrou (Green)), we aim to identify variants in either the expression or sequence of relevant genes which downstream could explain the differences in the ability of these kiwifruit strains to degrade protein.

We will -

Identify relevant publically available RNA-seq data sets relevant to the study.

Map these reads to the genome available from the [Kiwifruit Genome Database](https://kiwifruitgenome.org/) (*Actinidia chinensis* (Hong Yang v3)).

Identify gene expression variations between the different strains.

Identify gene sequence variations between the strains.

### Previously....

##### Collecting Read Data

Reads come in fastq files. These contain nucleotide (DNA or RNA depending) and quality information. They also come gziped (.gz suffix; compression format) as the files themselves can be massive. Thankfully, we do not have to uncompress the files as most bioinformatic tools work just as well on compressed and uncompressed files. There are many depositories of read data. Heavy-hitters include the [European Nucleotide Archive](https://www.ebi.ac.uk/ena/browser/) (ENA) and the [Sequence Read Archive](https://www.ncbi.nlm.nih.gov/sra/?term=) (SRA). The SRA (in my opinion) has an intuitive advanced searching system as it is connected to NCBI, but accessing reads from ENA is easier as there is an somewhat confusing system to downloading SRA files. Most projects are stored on both SRA and ENA.

We are trying to identify variants between strains under all conditions therefore we need a dataset that -

1)  Contains all the strains of interest.

2)  Samples under normal conditions.

SRA and ENA contains the reads, but the whole dataset (project) that the reads were generate through can be found through [Bioproject](https://www.ncbi.nlm.nih.gov/bioproject/).

Through searching for -

"Actinidia chinensis"[Organism] AND ("bioproject sra"[Filter] AND "method sequencing"[Filter]) AND "scope multispecies"[Filter]

on Bioproject I identified the project below.

| Title | Accession | Strains | Conditions |
|------------------|------------------|------------------|------------------|
| Multi-omics reveals the mechanisms underlying the differences in the formation of volatile and non-volatile metabolites of different kiwifruit varieties | PRJNA1413911 | G3, Hongyang, Lvrou (Red, Green and Yellow) | Normal |

Copying the accession into ENA takes you to the project page.

```{r, echo = F}
knitr::include_url("https://www.ebi.ac.uk/ena/browser/view/PRJNA1413911")
```

Clicking "Download All", downloads a file, which can be run through the university computer system to download all .fastq files associated with the project.

These are raw fastq files containing all reads collected. These reads could have all sorts of issues associated with them including -

-   Duplication

-   Still containing adapters (sequences that allow the sequencer (the machine reading the nucleotides) to be able to identify which nucleotides they are to read)

-   Truncated

Using read quality control tools such as [fastp](#0), we can identify and remove these from the set.

Afterwards, we need to map the reads to a genome. [Hisat](#0), is a commonly used tool which maps fastq files to a given genome creating commonly used file format - sequence aligment map (SAM). Sam files are a notoriously large, hard to work with file type. They are often converted to binary alignment map (BAM) files for downstream analysis using [Samtools](https://samtools.github.io/).

The bam files contain both sequence alignment information (where each read maps to on the genome) but also quantity information (how many reads map to each region of the genome). Therefore, using the bam files we can create a count-table that contains each transcript (RNA molecule) and how many times that transcript is seen across the different conditions. This is completed using [featureCounts](#0).

However, we also need to supply an annotation file - usually a GTF or GFF filer. This is supplied with the genome and tells featureCounts were each gene starts and ends.

Breakdown of steps completed before the lab

| Stage of Analysis |     Step 1     |     Step 2      | Step 3  |       Step 4       |
|:-------------:|:-------------:|:-------------:|:-------------:|:-------------:|
|     **Step**      | Download Files | Quality Control | Mapping | Create count table |
|     **Tools**     |    SRA/ENA     |      Fastp      |  Hisat  |   FeatureCounts    |

## Differential Expression Analysis

For identifying genes that are expressed at different levels between strains, we will use the tool [DESeq2](#0).

For this we need a count table and a meta table. The count table (created by featureCounts) contain the amount of times a transcript was mapped to a particular gene.

All required input files are supplied, put them all into the same folder and set that folder as your working directory (from bar above, Session, Set Working Directory, Choose Directory)

```{r}
# Read in Count Table 

kiwifruit_counts <- read.table("kiwifruit_counts_all2.txt")
colnames(kiwifruit_counts) <- c("HY_1", "HY_2", "HY_3", "G3_1", "G3_2", "G3_3", "LV_1", "LV_2", "LV_3") # Name then to what we can understnd. 
kiwifruit_counts <- kiwifruit_counts[rowSums(kiwifruit_counts) > 0, ] # Removing any rows where there are no counts for the transcrip. 
head(kiwifruit_counts)
```

A metatable tells DESeq2 all the important information about the types of samples in the set. It is important when we are comparing between conditions as it tells DESeq2 which columns are replicates and which are entirely different conditions.

```{r}
sample_names <- c("HY_1", "HY_2", "HY_3", "G3_1", "G3_2", "G3_3", "LV_1", "LV_2", "LV_3") # These need to match the colnames in the count_table. 

meta_table <- data.frame(SampleID = sample_names, Strain = c(rep("Hongyang", 3), rep("G3", 3), rep("Lvrou", 3))) # Which samples are which strain. 

rownames(meta_table) <- meta_table$SampleID

```

### Data Checking

We need to make sure that none of the samples are skewing our results. The first is to check whether there are any samples with significantly more or less reads. Below are two ways to do that. The first is looking directly at the amount of reads present, and the second is to look at the correlation of reads between samples.

For example, gene 1 has on on average 4 reads in species 1, and 8 reads in species 2. Gene 2 has 10 reads in species 1 and 2 in species 2. The different species have a different amount of reads in different genes. We can see which samples have similar amounts of reads for the same genes in the heat map below.

```{r}

colSums(kiwifruit_counts) / 1e6

log_counts <- log2(kiwifruit_counts + 1)

# Calculate sample-to-sample correlations
cor_matrix <- cor(log_counts, method = "pearson")
print(pheatmap(cor_matrix, annotation_col = meta_table, drop = FALSE))

```

**Question**

Are the samples part of the same species grouping together? Are there any samples with a lot less or a lot more reads than another?

Now we have a count_table a metatable, and have checked our data, we can move forward with DESeq2.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = kiwifruit_counts,
  colData = meta_table,
  design = ~ Strain # Group and compare samples across the different strains
)

dds$Strain <- relevel(dds$Strain, ref = "Hongyang") # DESeq automatically makes the "base" level through alphabetical order. As the genome that I mapped the reads to is the Hongyang strain, it is logical to compare the other strains, to the "reference". 

dds <- DESeq(dds) # Run DESeq
resultsNames(dds) # Shows you both strains compared to Hongyang. 
```

### Post-DESeq2 QC

```{r}
###############################################
#VST for PCA + QC
###############################################
vsd <- vst(dds, blind = FALSE)

###############################################
#PCA plot
###############################################
pca <- plotPCA(vsd, intgroup = "Strain") +
  ggplot2::ggtitle("PCA of Samples: Strains") +
  ggplot2::theme_minimal(base_size = 14)
pca

###############################################
#Sample distance heatmap
###############################################
dists <- dist(t(assay(vsd)))
mat   <- as.matrix(dists)

pheatmap(mat,
         clustering_distance_rows = dists,
         clustering_distance_cols = dists,
         main = "Sample-to-Sample Distance Heatmap" , annotation_col = meta_table, drop = FALSE, 
         fontsize = 12)

```

**Question**

Is there any sample that is not following the same pattern as its replicates?

```{r}

# Find summaries. Play around with changing the alpha value and the lncThreshold value and run different pathways analysis through them. See how the outputs change depending on what values you choose. 

res_G3_vs_HY <- results(dds, alpha = 0.05, lfcThreshold = 5, 
  contrast = list("Strain_G3_vs_Hongyang")) 

# Remembering back to the lectures, the alpha value tells DESeq the threshold for the false discovery rate (FDR), a value of 0.01 indicates that 1 % of the reported genes are false positives. 


res_LV_vs_HY <- results(dds, alpha = 0.05, lfcThreshold = 5,
  contrast = list("Strain_Lvrou_vs_Hongyang"))

summary(res_G3_vs_HY)
summary(res_LV_vs_HY)
```

**Question**

Based on above, which kiwifruit species are more similar than the other?

So, technically right now, differential expression analysis is done. But now we have to look into the results and see which genes are undergoing differential expression.

We can identify which genes are undergoing the greatest positive (upregulated) and negative (downregulated) logfold change (LFC).

```{r}

res_G3_vs_HY_df <- as.data.frame(res_G3_vs_HY)
res_LV_vs_HY_df <- as.data.frame(res_LV_vs_HY)

G3 <- res_G3_vs_HY_df %>% rownames_to_column("gene")
LV <- res_LV_vs_HY_df  %>% rownames_to_column("gene")

# 2. Top 5 up in G3
up_G3 <- G3 %>%
  filter(padj < 0.05, log2FoldChange > 0) %>%
  arrange(desc(log2FoldChange)) %>%
  head(5) %>%
  mutate(Comparison = "G3 vs HY", Direction = "Up")

# 3. Top 5 down in G3
down_G3 <- G3 %>%
  filter(padj < 0.05, log2FoldChange < 0) %>%
  arrange(log2FoldChange) %>%   # most negative first
  head(5) %>%
  mutate(Comparison = "G3 vs HY", Direction = "Down")

# 4. Top 5 up in LV
up_LV <- LV %>%
  filter(padj < 0.05, log2FoldChange > 0) %>%
  arrange(desc(log2FoldChange)) %>%
  head(5) %>%
  mutate(Comparison = "LV vs HY", Direction = "Up")

# 5. Top 5 down in LV
down_LV <- LV %>%
  filter(padj < 0.05, log2FoldChange < 0) %>%
  arrange(log2FoldChange) %>%
  head(5) %>%
  mutate(Comparison = "LV vs HY", Direction = "Down")

# 6. Combine everything
final_table <- bind_rows(up_G3, down_G3, up_LV, down_LV)

# 7. Pick only the columns you want to show
final_table <- final_table %>%
  dplyr::select(Comparison, Direction, gene, log2FoldChange, padj)

# 8. View it
final_table %>%
  mutate(
    log2FoldChange = round(log2FoldChange, 2),
    padj = format(padj, scientific = TRUE, digits = 3)
  ) %>%
  kable(
    col.names = c("Comparison", "Direction", "Gene", "log₂ Fold Change", "Adj. p‑value"),
    caption = "Top 5 up‑ and down‑regulated genes in each comparison",
    digits = 2,
    format = "html"  # or "markdown" for plain text
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

But this means nothing in of itself. We have to first annotate the genes. This "translates" the gene so that we can understand their function. This has been completed earlier using a tool called [eggnog](#0).

```{r}
# Read in annotation file
eggnog <- read_tsv(
  "eggnog_annotation.emapper.annotations",
  skip = 3,
  comment = "##",
  col_names = TRUE
)
names(eggnog)[1] <- "query"

# Map to genes within the DESeq2 result. 

eggnog <- eggnog %>%
  mutate(gene = sub("\\.t\\d+$", "", query))

res_all <- results(dds)
res_all_df <- as.data.frame(res_all)
res_all_df$gene <- rownames(res_all_df)
annotated_all <- merge(res_all_df, eggnog, by = "gene", all.x = TRUE)
```

Having gene annotations opens up the possibilities of what we can do with RNA-seq analysis.

For example, pathway expression analysis.

```{r}
# 1. Function to split comma‑separated GO terms
split_go <- function(go_string) {
  if (is.na(go_string) || go_string == "-" || go_string == "") {
    return(character(0))
  } else {
    terms <- trimws(unlist(strsplit(go_string, ",")))
    terms[terms != ""]
  }
}

# 2. Build gene‑to‑GO mapping in long format
gene2go <- eggnog[, c("gene", "GOs")]
gene2go <- gene2go[!is.na(gene2go$GOs) & gene2go$GOs != "-", ]

gene2go_long <- gene2go %>%
  rowwise() %>%
  mutate(GO = list(split_go(GOs))) %>%
  unnest(GO) %>%
  dplyr::select(gene, GO) %>%
  distinct()

# 3. Create term2gene and term2name for clusterProfiler
term2gene <- gene2go_long[, c("GO", "gene")]

go_ids <- unique(gene2go_long$GO)
go_names <- mapIds(GO.db, keys = go_ids, column = "TERM", keytype = "GOID")
term2name <- data.frame(
  ID = go_ids,
  Description = ifelse(is.na(go_names[go_ids]), go_ids, go_names[go_ids]),
  stringsAsFactors = FALSE
)

# 4. Define background genes (all genes in your DESeq2 object)
background_genes <- rownames(dds)   # if dds exists


# G3
up_genes_G3 <- G3 %>%
  filter(padj < 0.05, log2FoldChange > 0) %>%
  pull(gene)

down_genes_G3 <- G3 %>%
  filter(padj < 0.05, log2FoldChange < 0) %>%
  pull(gene)

# LV
up_genes_LV <- LV %>%
  filter(padj < 0.05, log2FoldChange > 0) %>%
  pull(gene)

down_genes_LV <- LV %>%
  filter(padj < 0.05, log2FoldChange < 0) %>%
  pull(gene)

term2gene <- gene2go_long[, c("GO", "gene")]

go_ids <- unique(gene2go_long$GO)
go_names <- mapIds(GO.db, keys = go_ids, column = "TERM", keytype = "GOID")
term2name <- data.frame(
  ID = go_ids,
  Description = ifelse(is.na(go_names[go_ids]), go_ids, go_names[go_ids]),
  stringsAsFactors = FALSE
)

# ---------- 2. Run enrichment for all 4 sets ----------
run_go <- function(gene_list, comp_name, dir_name) {
  if (length(gene_list) < 5) {
    message(paste("Skipping", comp_name, dir_name, "- too few genes"))
    return(NULL)
  }
  
  res <- enricher(
    gene = gene_list,
    TERM2GENE = term2gene,
    TERM2NAME = term2name,
    universe = background_genes,
    pvalueCutoff = 0.05,
    pAdjustMethod = "BH",
    minGSSize = 5,
    maxGSSize = 500
  )
  
  if (is.null(res) || nrow(res@result) == 0) {
    message(paste("No enrichment for", comp_name, dir_name))
    return(NULL)
  }
  
  # Add comparison labels to the results
  df <- res@result
  df$Comparison <- comp_name
  df$Direction <- dir_name
  return(df)
}

# Run all four
go_G3_up   <- run_go(up_genes_G3,   "G3 vs HY", "Up")
go_G3_down <- run_go(down_genes_G3, "G3 vs HY", "Down")
go_LV_up   <- run_go(up_genes_LV,   "LV vs HY", "Up")
go_LV_down <- run_go(down_genes_LV, "LV vs HY", "Down")

# ---------- 3. Combine top 5 GO terms from each into ONE table ----------
top_go_table %>%
  mutate(
    # Split GeneRatio into numerator and denominator
    num = as.numeric(sub("/.*$", "", GeneRatio)),
    den = as.numeric(sub("^.*/", "", GeneRatio)),
    GeneRatio = round(num / den, 3)
  ) %>%
  dplyr::select(-num, -den) %>%
  mutate(
    p.adjust = format(as.numeric(p.adjust), scientific = TRUE, digits = 3),
    geneID = gsub("/", ", ", geneID)   # base R alternative to str_replace_all
  ) %>%
  kable(
    col.names = c("Comparison", "Direction", "GO ID", "Description", 
                  "Adj. p‑value", "Gene Ratio", "Count", "Genes"),
    caption = "Top 5 enriched GO terms per group",
    digits = 3
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

```{r}
# ---------- Figure: GO enrichment across all groups ----------

# Function to run enrichment and return a data frame with labels
get_go_df <- function(genes, comp, dir) {
  if (length(genes) < 5) return(NULL)
  res <- enricher(genes, TERM2GENE = term2gene, TERM2NAME = term2name,
                  universe = background_genes, pvalueCutoff = 0.05,
                  pAdjustMethod = "BH", minGSSize = 5, maxGSSize = 500)
  if (is.null(res) || nrow(res@result) == 0) return(NULL)
  df <- res@result
  df$Comparison <- comp
  df$Direction  <- dir
  return(df)
}

# Get all four
go_G3_up   <- get_go_df(up_genes_G3,   "G3 vs HY", "Up")
go_G3_down <- get_go_df(down_genes_G3, "G3 vs HY", "Down")
go_LV_up   <- get_go_df(up_genes_LV,   "LV vs HY", "Up")
go_LV_down <- get_go_df(down_genes_LV, "LV vs HY", "Down")

# Combine and keep top 10 per group
all_go <- bind_rows(go_G3_up, go_G3_down, go_LV_up, go_LV_down)
all_go$Group <- paste(all_go$Comparison, all_go$Direction, sep = "\n")

top_terms <- all_go %>%
  group_by(Group) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(Description = factor(Description, levels = rev(unique(Description[order(p.adjust)]))))

plot_group <- function(data, group_label) {
  df <- data %>% filter(Group == group_label)
  
  ggplot(df, aes(x = GeneRatio, y = Description)) +
    geom_point(aes(size = Count, color = p.adjust)) +
    scale_color_gradient(low = "red", high = "blue", trans = "log10",
                         name = "Adjusted p‑value") +
    scale_size_continuous(name = "Gene Count") +
    labs(x = "Gene Ratio", y = NULL,
         title = group_label) +
    theme_bw() +
    theme(axis.text.y = element_text(size = 10),
          plot.title = element_text(face = "bold", size = 14),
          panel.grid.minor = element_blank())
}
p_G3_up   <- plot_group(top_terms, "G3 vs HY\nUp")
p_G3_down <- plot_group(top_terms, "G3 vs HY\nDown")
p_LV_up   <- plot_group(top_terms, "LV vs HY\nUp")
p_LV_down <- plot_group(top_terms, "LV vs HY\nDown")

print(p_G3_down)
print(p_G3_up)
print(p_LV_up)
print(p_LV_down)
```

**Question**

1)  Which pathways are expressed differently between the different strains and what might they mean regarding kiwifruit physiology.

We are trying to identify the cause of why green kiwifruit have a greater ability to digest proteins. The type of protein that degrades protein is called a protease. Using the annotation file, we can identify which genes contain a protease domain. Any genetic (sequential or expression) differences between the strains, could be the answer as to why the phenotype is different.

```{r}
# Define keywords for proteases (case‑insensitive)
protease_keywords <- c("protease", "peptidase", "proteinase", 
                       "endopeptidase", "exopeptidase", "carboxypeptidase",
                       "aminopeptidase", "dipeptidase", "metalloprotease",
                       "serine protease", "cysteine protease", "aspartic protease")

# Filter rows where Description or GOs contain any keyword
protease_genes <- eggnog %>%
  filter(
    str_detect(Description, regex(paste(protease_keywords, collapse = "|"), ignore_case = TRUE)) |
    str_detect(GOs, regex(paste(protease_keywords, collapse = "|"), ignore_case = TRUE))
  ) %>%
  pull(gene) %>%          # extract gene IDs
  unique()

cat("Number of putative protease genes:", length(protease_genes), "\n")

# Map the proteases to the expression data. This allows us to identify which proteases are changing in expression between the strains. 

# Create data frames for each comparison with gene and statistics
G3_expr <- G3 %>% filter(gene %in% protease_genes)   
LV_expr <- LV %>% filter(gene %in% protease_genes)

# For each, classify as up/down if padj < 0.05 and |log2FC| > 2 (your threshold)
G3_proteases <- G3_expr %>%
  mutate(Regulation = case_when(
    padj < 0.05 & log2FoldChange > 2   ~ "Up in G3",
    padj < 0.05 & log2FoldChange < -2  ~ "Down in G3",
    TRUE ~ "Not significant"
  ))

LV_proteases <- LV_expr %>%
  mutate(Regulation = case_when(
    padj < 0.05 & log2FoldChange > 2   ~ "Up in LV",
    padj < 0.05 & log2FoldChange < -2  ~ "Down in LV",
    TRUE ~ "Not significant"
  ))

# Combine into one table for easy viewing
protease_table <- G3_proteases %>%
  dplyr::select(gene, log2FC_G3 = log2FoldChange, padj_G3 = padj, Reg_G3 = Regulation) %>%
  full_join(
    LV_proteases %>% 
      dplyr::select(gene, log2FC_LV = log2FoldChange, padj_LV = padj, Reg_LV = Regulation),
    by = "gene"
  )

# Add annotation description (optional)
protease_table <- protease_table %>%
  left_join(eggnog %>% dplyr::select(gene, Description), by = "gene") %>%
  relocate(Description, .after = gene)

# Show only significant ones (either comparison)
protease_table_sig %>%
  mutate(
    log2FC_G3 = round(log2FC_G3, 2),
    log2FC_LV = round(log2FC_LV, 2),
    padj_G3 = format(padj_G3, scientific = TRUE, digits = 2),
    padj_LV = format(padj_LV, scientific = TRUE, digits = 2)
  ) %>%
  dplyr::select(Gene = gene, Description, 
                `log2FC (G3)` = log2FC_G3, `p‑adj (G3)` = padj_G3, `Reg (G3)` = Reg_G3,
                `log2FC (LV)` = log2FC_LV, `p‑adj (LV)` = padj_LV, `Reg (LV)` = Reg_LV) %>%
  kable(caption = "Significant protease genes") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  print()
```

```{r}
cat("Significant proteases in G3 vs HY:\n")
table(protease_table_sig$Reg_G3)

cat("\nSignificant proteases in LV vs HY:\n")
table(protease_table_sig$Reg_LV)
```

```{r}
# Significant protease genes 
sig_protease_genes <- protease_table_sig$gene 

# Get expression counts for proteases from your original count table

protease_counts <- kiwifruit_counts[rownames(kiwifruit_counts) %in% sig_protease_genes, ]

# Optional: log2 transform and scale per gene
log_counts <- log2(protease_counts + 1)
scaled <- t(scale(t(log_counts)))   # z‑score per gene

# Annotate columns by strain
annotation_col <- data.frame(Strain = meta_table$Strain)
rownames(annotation_col) <- colnames(kiwifruit_counts)

pheatmap(scaled,
         annotation_col = annotation_col,
         show_rownames = FALSE,   # too many genes; can show if few
         main = "Protease gene expression across strains",
         fontsize = 8)
```

**Question**

1)  How many protease genes are changing in expression across strains. Remember which colour kiwifruit has the greatest protease ability.

Actinidian belongs to the peptidase C1 family. How many C1's are there in the genome?

```{r}
cysteine_protease_keywords <-  "peptidase C1 family"

cysteine_protease_genes <- eggnog %>%
  filter(
    str_detect(Description, regex(paste(cysteine_protease_keywords, collapse = "|"), ignore_case = TRUE)) |
    str_detect(GOs, regex(paste(cysteine_protease_keywords, collapse = "|"), ignore_case = TRUE))
  ) %>%
  pull(gene) %>%          # extract gene IDs
  unique()

cat("Number of putative protease genes:", length(cysteine_protease_genes), "\n")

# Map the proteases to the expression data. This allows us to identify which proteases are changing in expression between the strains. 

# Create data frames for each comparison with gene and statistics
G3_expr <- G3 %>% filter(gene %in% cysteine_protease_genes)   
LV_expr <- LV %>% filter(gene %in% cysteine_protease_genes)

# For each, classify as up/down if padj < 0.05 and |log2FC| > 2 (your threshold)
G3_proteases_cysteine <- G3_expr %>%
  mutate(Regulation = case_when(
    padj < 0.05 & log2FoldChange > 2   ~ "Up in G3",
    padj < 0.05 & log2FoldChange < -2  ~ "Down in G3",
    TRUE ~ "Not significant"
  ))

LV_proteases_cysteine <- LV_expr %>%
  mutate(Regulation = case_when(
    padj < 0.05 & log2FoldChange > 2   ~ "Up in LV",
    padj < 0.05 & log2FoldChange < -2  ~ "Down in LV",
    TRUE ~ "Not significant"
  ))

# Combine into one table for easy viewing
cysteine_protease_table <- G3_proteases_cysteine %>%
  dplyr::select(gene, log2FC_G3 = log2FoldChange, padj_G3 = padj, Reg_G3 = Regulation) %>%
  full_join(
    LV_proteases_cysteine %>% 
      dplyr::select(gene, log2FC_LV = log2FoldChange, padj_LV = padj, Reg_LV = Regulation),
    by = "gene"
  )

# Add annotation description (optional)
cysteine_protease_table <- cysteine_protease_table %>%
  left_join(eggnog %>% dplyr::select(gene, Description), by = "gene") %>%
  relocate(Description, .after = gene)

# Show only significant ones (either comparison)
cysteine_protease_table_sig <- cysteine_protease_table %>%
  filter(Reg_G3 != "Not significant" | Reg_LV != "Not significant")

cysteine_protease_table_sig %>%
  mutate(
    log2FC_G3 = round(log2FC_G3, 2),
    log2FC_LV = round(log2FC_LV, 2),
    padj_G3 = format(padj_G3, scientific = TRUE, digits = 2),
    padj_LV = format(padj_LV, scientific = TRUE, digits = 2)
  ) %>%
  dplyr::select(Gene = gene, Description, 
                `log2FC (G3)` = log2FC_G3, `p‑adj (G3)` = padj_G3, `Reg (G3)` = Reg_G3,
                `log2FC (LV)` = log2FC_LV, `p‑adj (LV)` = padj_LV, `Reg (LV)` = Reg_LV) %>%
  kable(caption = "Significant cysteine protease genes (C1 family)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  print()
```

```{r}
# Significant protease genes 
cystenine_sig_protease_genes <- cysteine_protease_table_sig$gene 

# Get expression counts for proteases from your original count table

cysteine_protease_counts <- kiwifruit_counts[rownames(kiwifruit_counts) %in% cystenine_sig_protease_genes, ]

# Optional: log2 transform and scale per gene
cysteine_log_counts <- log2(cysteine_protease_counts + 1)
cysteine_scaled <- t(scale(t(cysteine_log_counts)))   # z‑score per gene

# Annotate columns by strain
annotation_col <- data.frame(Strain = meta_table$Strain)
rownames(annotation_col) <- colnames(kiwifruit_counts)

pheatmap(cysteine_scaled,
         annotation_col = annotation_col,
         show_rownames = FALSE,   # too many genes; can show if few
         main = "Cysteine protease gene expression across strains",
         fontsize = 8)
```

**Question**

Ignoring any that are non-significant in any condition, how many proteases / peptidases are we left with? Is it likely that the difference in phenotype is due to the changes in gene expression that we are / are not seeing?
